Demography and linked selection in wild and domesticated maize

### 2015-10-18 manuscript.tex
updates

\significancetext{Both selection and demographic change play important roles in shaping diversity across the genome, but clear empirical examples of the interplay of these two forces are lacking.
Here we document the combined effects of demography and linked selection on genome-wide diversity in domesticated maize and its wild ancestor teosinte.  We estimate that maize underwent a bottleneck of 5\% of the size of the ancestral teosinte population, but that recent expansion has resulted in a maize population perhaps orders of magnitude larger than teosinte.  We show that positive selection on new mutations has had relatively little effect on genetic diversity, but that selection against deleterious mutations has dramatically reduced diversity in and immediately around genes in both taxa.  We find that the relative effect of selection depend qualitatively on the age of the polymorphisms evaluated: while older polymorphisms in maize show more limited effects of linked selection, new mutations instead reflect its larger current size and more efficient purifying selection. Our results demonstrate that a complete understanding of genome-wide patterns of diversity will require careful assessment of both demographic history and the effects of linked selection. }

@@

\begin{abstract}
  Unique selective and demographic forces operate on domesticated plant species. \mbh{I don't agree with this statement; characteristics of demography and selection vary across crops and I wouldn't be surprised to find similar demographies and patterns of selection between crop and wild species} These forces interact during and after domestication to generate the patterns of DNA variability that are persistent today. To quantify the interplay between demography and selection, we investigated genetic diversity in maize, one of the most important crops for food, feed, and fuel world-wide. Our sample included whole genome sequence data from 23 maize and 13 teosinte individuals.  We obtained a complete estimate of the population size fluctuations and other demographic parameters experienced by maize as it was domesticated from teosinte. Here, we show that maize went through a domestication bottleneck with a population size of approximately 5\% that of teosinte before it experienced rapid population size expansion post-domestication. We observe that hard sweeps on genic mutations are not the primary force driving maize evolution. We find that a reduced population size during domestication decreased the efficiency of purifying selection to purge deleterious alleles from maize. However, expansion after domestication has since increased the efficiency of purifying selection to levels exceeding those seen in teosinte. Our results demonstrate that particularly in domesticated species or bottlenecked species \mbh{right, observations can be generalized to wild species as well; I don't think there's any need to say crops are a special case}, demographic and selective history in the ancient and recent past both contribute to genetic variability that is present today, providing substantial implications for the continued improvement of domesticated species. \mbh{I'd end this more generally, expressing the importance of knowing demography when characterizing selection genome-wide}

@@

\dropcap{T}he genetic properties of populations are determined by a constant interplay between genetic drift and selection. Drift is the consequence of a finite population size, such that random sampling of gametes every generation will incrementally lead to genetic changes \cite{dobzhansky1957}. In other words, the demographic history of a population, which includes its population size fluctuations and mating patterns, has a long-lasting impact on the patterns of DNA variability exhibited by that population. Conversely, selection systematically alters a population by favoring particular alleles at the expense of others. In contrast to drift, which results from demographic history, selection is the result of the relationship between genotypes, phenoytpes, and ultimately fitness. Importantly, demography and selection do not operate independently to determine genetic variability \cite{li2012}. Instead, these forces are engaged in a continuous battle by which each masks and obfuscates the outcomes of the other, ultimately meaning that the selective history of a species cannot be properly evaluated without an thorough estimate of demographic history, and demography cannot be evaluated without an understanding of selection.

Therefore, to estimate the historical demographic parameters of a population, researchers often utilize supposed neutrally-evolving DNA  \cite{luikart2003, gutenkunst2009}. However, it has been called into question whether reasonable approaches exist
for identifying neutral regions of the genome, since the effects of
selection can be wide-ranging \cite{li2012, slotte2014}. For example, in
\emph{Drosophila} it appears that the majority of the genome is
impacted by the effects of selection through linkage
\cite{sella2009}. One technique that has been utilized to estimate demographic parameters independently of selective parameters is to limit analyses to only sites far from genes
\cite{gazave2014}, but as \emph{Drosophila} demonstrates, it is
difficult to be certain, without species-specific knowledge, that sites even in gene-poor regions of the
genome are not influenced by linked selection. Ultimately, an understanding
of which sites have the potential to be adaptive, and how these affect
genome-wide patterns of variability through linkage, is required for
reasonable demographic inference.

Maize represents an excellent organism to study these phenomena. Maize is a species of tremendous importance worldwide as both a staple crop \cite{shiferaw2011} and as a model for understanding plant evolution \cite{strable2009}. 
Archaeological and genetic studies have
established that maize domestication is likely to have taken place in
 Mexico approximately 9,000 years bp
\cite{smith1995,matsuoka2002}. Teosinte, the most
recent wild ancestor to maize, remains extant throughout much of the
Americas \cite{wilkes1967}. Additionally, several large-effect
domestication loci \cite{doebley1995, wills2013, wang2015} and putative domestication
regions \cite{hufford2012} have been identified. But despite all that is
known about maize domestication, the parameters of the
domestication process remain uncertain. Specifically, the size of the
maize domestication bottleneck has not been estimated independently of
the bottleneck's duration, nor are there sequence-based estimates of the effective
population size of modern maize. At the same time, the predominant form or forms of selection that have taken place during and since domestication remain unknown. Sequence information from maize and teosinte may therefore be utilized to address these questions.

@@

\begin{figure*}[!tb]
%\vspace*{.05in}
\centering
\includegraphics[width=.4\textwidth]{FigsAndFiles/DomesticationModel/domesticationModel.pdf}
\caption{Parameter estimates for a simple bottleneck model of maize domestication. See methods for details. \label{fig:bottleneck} }
\end{figure*}

@@

Our findings demonstrate that classic hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize, a result we show is robust to concerns about power due to the effects of background selection  \cite{enard2014}. 
Although our approach ignores the potential for hard sweeps in noncoding regions of the genome, a growing body of evidence argues against hard sweeps as the prevalent mode of selection shaping maize variability. 
Among well-characterized domestication loci, only the gene \emph{tga1} shows evidence of a hard sweep on a missense mutation \cite{wang2015}, while several loci are consistent with ``soft sweeps''  from standing variation \cite{studer2011,gallavotti2004role} or multiple mutations \cite{wills2013}. 
Moreover, genome-wide studies of domestication \cite{hufford2012},  local adaptation \cite{Takuno15062015} and modern breeding \cite{beissinger2014} all support the importance of standing variation and/or non-coding variants as primary sources of adaptive variation. 
Soft sweeps are expected to be common when $2N_e\mu_b \ge 1$, where $\mu_b$ is the mutation rate of beneficial alleles with selection coefficient $s_b$ \cite{messer2013population}.  \tmb{Jeff, you put this here, but as written I don't think we need to mention $s_b$}
Assuming a mutation rate of $3 \times 10^{-8}$ \cite{clark2005} and that on the order of $\approx 1-5\%$ of mutations are beneficial \cite{eyre2007distribution}, this implies that soft sweeps should be common in both maize and teosinte for mutational targets $>>10kb$ --- a plausible size for quantitative traits or for regulatory evolution targeting genes with large up- or down-stream control regions \cite[e.g.]{studer2011}.
Indeed, many adaptive traits in both maize \cite{wallace2014} and teosinte \cite{weber2008} are highly quantitative,  and adaptation in both maize \cite{hufford2012} and teosinte \cite{pyhajarvi2013complex} has involved selection on regulatory variation.

@@

A consequence of the previously-inefficient purifying selection in maize is that it likely harbors more weakly deleterious alleles segregating at intermediate frequency than does teosinte. Therefore, high genetic load could be a part of the explanation of why maize inbreds have continued to improve over the past several decades \cite{meghji1984}. Our results also demonstrate that recent purifying selection in maize has become much more effective, potentially explaining the ongoing improvement of these inbreds as maize lines are continuously selected. Additionally, the large $N_e$ of modern maize compared to modern teosinte implies that for new mutations, selection will operate much more efficiently in maize. This observation is not restricted to maize. Bottlenecks are have been documented or predicted across most domesticated plant and animal species \cite{doebley2006, hyten2006, zhu2007, bovine2009, larson2013 }, but rapid expansion post-bottleneck, which we demonstrate generates ubiquitous consequences relating to genetic diversity, is not well studied. In human, model, and other non-domesticated species, these results demonstrate that observed patterns of selection may depend heavily on recent and/or ancient patterns of demography.

@@

\subsection{Plant materials}
We made use of published sequences from inbred accessions of teosinte (Z. \emph{mays} ssp. \emph{parviglumis}) and maize landraces from the Maize HapMap3 panel as part of the Panzea project  \cite{chia2012, lemmon2014, hapmap3}. 
From these data, we removed 4 teosinte individuals that were not ssp. \emph{parviglumis} or appeared as outliers in an initial principal component analysis conducted with the package adegenet \cite{jombart2011} (Figure \ref{sFig:PCA}), leaving 13 teosinte and 23 maize that were used for all subsequent analyses (Table \ref{sTab:list}). We also utilized a single individual of (\emph{Tripsacum dactyloides}) as an outgroup.  All bam files are available at \textcolor{red}{/iplant/home/shared/panzea/hapmap3/bam\_internal/v3\_bams\_bwamem}). \jri{some clarification here. did we use Lemmon's higher depth teosinte? If so, why does Li say we only had 7 high-depth bam files?  shouldn't there be more?}

>>>>

\significancetext{Both selection and demographic change play important roles in shaping diversity across the genome, but clear empirical examples of the interplay of these two forces are lacking.
Here we document the combined effects of demography and linked selection on genome-wide diversity in domesticated maize and its wild ancestor teosinte.  We estimate that maize underwent a bottleneck of 5\% of the size of the ancestral teosinte population, but that recent expansion has resulted in a maize population perhaps orders of magnitude larger than teosinte.  We show that positive selection on new genic mutations has had relatively little effect on genetic diversity, but that selection against deleterious mutations has dramatically reduced diversity in and immediately around genes in both taxa.  We find that the relative effect of selection depend qualitatively on the age of the polymorphisms evaluated: while older polymorphisms in maize show more limited effects of linked selection, new mutations instead reflect its larger current size and more efficient purifying selection. Our results demonstrate that a complete understanding of genome-wide patterns of diversity will require careful assessment of both demographic history and the effects of linked selection. }

@@

\begin{abstract}
  The genetic variability present in a population is increases through mutation and drift, and generally reduced via selection. But these forces do not act in a vacuum, and instead interact continuously to shape patterns of genetic diversity. To quantify the interplay between demography (which drives drift) and selection, we investigated genetic diversity in maize, one of the most important crops for food, feed, and fuel world-wide. Our sample included whole genome sequence data from 23 maize and 13 teosinte individuals.  We obtained a complete estimate of the population size fluctuations and other demographic parameters experienced by maize as it was domesticated from teosinte. Here, we show that maize went through a domestication bottleneck with a population size of approximately 5\% that of teosinte before it experienced rapid population size expansion post-domestication. We observe that hard sweeps on genic mutations are not the primary force driving maize evolution. We find that a reduced population size during domestication decreased the efficiency of purifying selection to purge deleterious alleles from maize. However, expansion after domestication has since increased the efficiency of purifying selection to levels exceeding those seen in teosinte. Our results demonstrate that demographic and selective history in the ancient and recent past both contribute to the levels of genetic variability that are present today. Our results also show that for bottlenecked species, rapid post-bottleneck expansion can have wide-ranging impacts on diversity that conflict with the expected loss of diversity that occurs during the bottleneck itself.

@@

\dropcap{T}he genetic properties of populations are determined by a constant interplay between genetic drift and selection. Drift is the consequence of a finite population size, such that random sampling of gametes every generation will incrementally lead to genetic changes \cite{dobzhansky1957}. In other words, the demographic history of a population, which includes its population size fluctuations and mating patterns, has a long-lasting impact on the patterns of DNA variability displayed by that population. Conversely, selection systematically alters a population by favoring particular alleles at the expense of others. In contrast to drift, which results from demographic history, selection is the result of the relationship between genotypes, phenoytpes, and ultimately fitness. Importantly, demography and selection do not operate independently to determine genetic variability \cite{li2012}. Instead, these forces are engaged in a continuous battle by which each masks and obfuscates the outcomes of the other, ultimately meaning that the selective history of a species cannot be properly evaluated without a thorough estimate of demographic history, and demography cannot be evaluated without an understanding of selection.

For instance, to estimate the historical demographic parameters of a population, researchers often utilize supposed neutrally-evolving DNA  \cite{luikart2003, gutenkunst2009}. However, it has been called into question whether reasonable approaches exist
for identifying neutral regions of the genome, since the effects of
selection can be wide-ranging \cite{li2012, slotte2014}. In \emph{Drosophila}, for example, the majority of the genome is
impacted by the effects of selection through linkage
\cite{sella2009}. One technique that has been utilized to estimate demographic parameters independently of selective parameters is to limit analyses to only sites far from genes
\cite{gazave2014}, but as \emph{Drosophila} demonstrates, it is
difficult to be certain, without species-specific knowledge, that sites even in gene-poor regions of the
genome are not influenced by linked selection. Ultimately, an understanding
of which sites have the potential to be adaptive, and how these affect
genome-wide patterns of variability through linkage, is required for
reasonable demographic inference. At the same time, since the effects of demography alter genetic variability genome-wide \cite{akey2009}, and because the efficiency of selection depends on demographic characteristics such as population size \cite{kimura1963}, selection is equally difficult to quantify without properly accounting for demographic history.

Maize (\emph{Zea mays}) represents an excellent organism to study this interplay. Maize is a species of tremendous importance worldwide as both a staple crop \cite{shiferaw2011} and as a model for understanding plant evolution \cite{strable2009}. 
Archaeological and genetic studies have
established that maize domestication is likely to have taken place in
 Mexico approximately 9,000 years bp
\cite{smith1995,matsuoka2002}. Teosinte, the most
recent wild ancestor to maize, remains extant throughout much of the
Americas \cite{wilkes1967}. Additionally, several large-effect
domestication loci \cite{doebley1995, wills2013, wang2015} and putative domestication
regions \cite{hufford2012} have been identified. But despite all that is
known about maize domestication, the demographic parameters of the
domestication process remain uncertain. Specifically, the size of the
maize domestication bottleneck has not been estimated independently of
the bottleneck's duration, nor are there sequence-based estimates of the effective
population size of modern maize. At the same time, the predominant form or forms of selection that have taken place during and since domestication remain unknown. Sequence information from maize and teosinte may therefore be utilized to address these questions, and to better explain the interaction between selection and demography in a system with complex but quantifiable selective and demographic histories.

@@

\begin{figure*}[!tb]
%\vspace*{.05in}
\centering
\includegraphics[width=.4\textwidth]{FigsAndFiles/DomesticationModel/domesticationModel.pdf}
\caption{Parameter estimates for a basic bottleneck model of maize domestication. See methods for details. \label{fig:bottleneck} }
\end{figure*}

@@

Our findings demonstrate that classic hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize, a result we show is robust to concerns about power due to the effects of background selection  \cite{enard2014}. 
Although our approach ignores the potential for hard sweeps in noncoding regions of the genome, a growing body of evidence argues against hard sweeps as the prevalent mode of selection shaping maize variability. 
Among well-characterized domestication loci, only the gene \emph{tga1} shows evidence of a hard sweep on a missense mutation \cite{wang2015}, while several loci are consistent with ``soft sweeps''  from standing variation \cite{studer2011,gallavotti2004role} or multiple mutations \cite{wills2013}. 
Moreover, genome-wide studies of domestication \cite{hufford2012},  local adaptation \cite{Takuno15062015} and modern breeding \cite{beissinger2014} all support the importance of standing variation and/or non-coding variants as primary sources of adaptive variation. 
Soft sweeps are expected to be common when $2N_e\mu_b \ge 1$, where $\mu_b$ is the mutation rate of beneficial alleles with selection coefficient $s_b$ \cite{messer2013population}.  \tmb{Jeff, rewrote this part, and I like your writing here, but as written I don't think we need to mention $s_b?$}
Assuming a mutation rate of $3 \times 10^{-8}$ \cite{clark2005} and that on the order of $\approx 1-5\%$ of mutations are beneficial \cite{eyre2007distribution}, this implies that soft sweeps should be common in both maize and teosinte for mutational targets $>>10kb$ --- a plausible size for quantitative traits or for regulatory evolution targeting genes with large up- or down-stream control regions \cite[e.g.]{studer2011}.
Indeed, many adaptive traits in both maize \cite{wallace2014} and teosinte \cite{weber2008} are highly quantitative,  and adaptation in both maize \cite{hufford2012} and teosinte \cite{pyhajarvi2013complex} has involved selection on regulatory variation.

@@

A consequence of the previously-inefficient purifying selection in maize is that it likely harbors more weakly deleterious alleles segregating at intermediate frequency than does teosinte. Therefore, high genetic load could be a part of the explanation of why maize inbreds have continued to improve over the past several decades \cite{meghji1984}. Our results also demonstrate that recent purifying selection in maize has become much more effective, potentially explaining the ongoing improvement of these inbreds as maize lines are continuously selected. Additionally, the large $N_e$ of modern maize compared to modern teosinte implies that for new mutations, selection will operate much more efficiently in maize. This observation is not restricted to maize. Bottlenecks are have been documented or predicted across most domesticated plant and animal species \cite{doebley2006, hyten2006, zhu2007, bovine2009, larson2013 }, but rapid expansion post-bottleneck, which we demonstrate generates ubiquitous consequences relating to genetic diversity, is not well studied. In human, model, and other species, these results demonstrate that observed patterns of selection depend heavily on recent and/or ancient patterns of demography.

@@

\subsection{Plant materials}
We made use of published sequences from inbred accessions of teosinte (Z. \emph{mays} ssp. \emph{parviglumis}) and maize landraces from the Maize HapMap3 panel as part of the Panzea project  \cite{chia2012, lemmon2014, hapmap3}. 
From these data, we removed 4 teosinte individuals that were not ssp. \emph{parviglumis} or appeared as outliers in an initial principal component analysis conducted with the package adegenet \cite{jombart2011} (Figure \ref{sFig:PCA}), leaving 13 teosinte and 23 maize that were used for all subsequent analyses (Table \ref{sTab:list}). We also utilized a single individual of (\emph{Tripsacum dactyloides}) as an outgroup.  All bam files are available at \textcolor{red}{/iplant/home/shared/panzea/hapmap3/bam\_internal/v3\_bams\_bwamem}).

%\jri{some clarification here. did we use Lemmon's higher depth teosinte? If so, why does Li say we only had 7 high-depth bam files?  shouldn't there be more?}

% TB Says: Most bam files were not high-depth, but a few were. I used what was made available from Panzea group via iplant in summer 2014, and precise files that were used are linked and described in table. ANGSD GL approach makes it OK that depth varied.
